### 2020-11-21 pldi21/case-studies.tex
Add basic_adjustments figure
updates

\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-1.png}
      \caption{}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-2.png}
      \caption{}
    \end{subfigure}
  \end{center}
  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
  \label{fig:img-transformation}
\end{figure*}

@@

Motivated by this interview, 
we prototyped a collection of photo filter livelits.
One of these, \li{\$something} is demonstrated in Fig.~\ref{fig:img-transformation},
within a function, \li{classic_look}, that serveas as a ``preset'' that is 
mapped over a list of images (loaded by URL) at the bottom of the figure. 
% We show two for space reasons.

...\todo{finalize?}{}
This livelit takes 
a URL to an image as a parameter, and contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}, e.g. a variable. 
The livelit shows a live preview of the transformed image.
The expansion generates calls to a browser image processing library, 
not shown.

Because the livelit appears inside a function applied (by \li{map}) twice, 
there are now two closures associated with the livelit. 
Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression via a simple sidebar toggle. 
This causes the view to change as shown in ...\todo{todo}{}
The live feedback is based on the selected closure.
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. 
The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

>>>>

\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-1.png}
      \caption{}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-2.png}
      \caption{}
    \end{subfigure}
  \end{center}
  \caption{Case Study II: Image Transformation. The image shown is determined based on the selected closure.}
  \label{fig:img-transformation}
\end{figure*}

@@

Motivated by this interview, 
we prototyped a collection of photo filter livelits.
One of these, \li{\$basic_adjustments}, is demonstrated in Fig.~\ref{fig:img-transformation}. 
This livelit contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}, e.g. a variable. 
The livelit shows a live preview of the transformed image.
The expansion generates calls to a browser image processing framework, 
not shown.

This livelit is used within a function, \li{classic_look}, that creates a ``preset'' filter. 
This function is mapped over a list of images (loaded by URL) at the bottom of the figure. 
% We show two for space reasons.
The provided URL is passed into to livelit as a parameter.

Because the livelit appears inside a function applied (by \li{map}) twice, 
there are now two closures associated with the livelit. 
Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression via a simple sidebar toggle,
shown in the middle of Fig.~\ref{fig:img-transformation}.
This causes the view to change as shown. 
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. 
The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

### 2020-11-21 pldi21/implementation.tex
updates

\section{Implementation}\label{sec:implementation}
Implementing livelits requires tight integration between a rich editor, 
a type checker, and a live evaluator. The evaluator is used
both for running the model-view-update-expand system that controls 
each livelit and for evaluating splices. Consequently, 
the evaluator must be capable of evaluating incomplete programs and gathering hole closures
as specified by Hazelnut Live \cite{HazelnutLive}.

\subsection{Hazel}
Hazelnut Live is the formal foundation for the Hazel programming environment, so 
it was natural to choose the the Hazel programming environment for our primary implementation. 
 Hazel is a browser-based live functional programming environment 
organized from the ground up around typed holes. Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous.

Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
The core of Hazel is written in a pure functional style, so our implementation is able to closely 
follow the formalism. The Hazel language has been evolving in parallel with our work, 
and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
to livelit definitions suitable for inclusion in the livelit context used in the semantics.

Hazel achieves its ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
while the user edits 
the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
although the developers are aiming to maintain a text-like programming experience when editing 
symbolic code (this effort is orthogonal to our own). 
This means that it is conceptually straightforward to hide the model and splice data and instead display 
the computed view. However, there are a number of non-trivial layout challenges that come up, which we discuss
in Sec.~\ref{sec:layout} below.

\subsection{Text Editor Integration}
Before delving into layout issues, let us note that livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept textual program editor, based on Sketch-n-Sketch \cite{sns-pldi,sns-uist}, with support for livelits.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} 
is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
there are critical errors.

\subsection{Layout}\label{sec:layout}
Whether implemented in a structure editor or a text editor, livelits present 
interesting layout challenges--it can be difficult to sensibly 
interleave symbolic code with livelit views in a clean, readable manner.

Hazel uses an optimizing pretty printer to determine layout, an approach 
based on the work of \citet{DBLP:journals/pacmpl/Bernardy17}. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
which must be exactly one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
and a specified number of code lines (we omitted these specifications in Fig.~\ref{fig:color-impl}). 
The pretty printer lays out multi-line livelits
on their own line, as seen in Fig.~\ref{fig:grading}.

We are currently exploring a more flexible system whereby 
multi-line livelits can be named and applied within a larger expression, but
the GUI is not displayed until the next line, even if there are intervening characters.
This is inspired by the \verb|~| placeholder for large textual literals in the 
Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
options, e.g. pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
on a separate canvas or document while still formally being located within an underlying functional program.

This latter option 
would be particularly interesting for end-user programming scenarios: users with limited
programming experience 
could interact with a collection of livelits laid out separately in the popular ``dashboard'' style, 
without necessarily
even being aware that their interactions are actually edits to an underlying typed
functional program. More experienced users on the team could then make programmatic use of the entered
data and parameters in a seamless manner. We are also excited about the possibility of end users
getting curious enough to ``View Source'' in this setting, and in so doing 
being exposed to typed functional programming concepts gradually.

The layout of splices within livelits is also an interesting problem, and our current approach, where 
the provider must specify a fixed size for each splice, is not particularly flexible. 
A more principled system for GUI layout that integrates with program pretty 
printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
a generalized formula bar in the editor user interface that multiple livelits could share.
While splices can in principle contain other livelits, it is difficult for small splice editors to 
visually accomodate large livelit views in practice. 
Fortunately, variables can be used to support composition via substitution 
rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.

The Hazel pretty printer uses memoization and incremental diffs to maintain reasonable performance,
and adding livelits to the system did not substantially interfere with the existing approach.
For the relatively simple livelits that we have developed, performance has been adequate.
However, more complex  livelits or livelits that update quickly 
could easily cause problems with editor responsiveness.
This could perhaps be resolved by performing livelit-related computations asynchronously. Care 
would need to be taken to ensure that intervening edits do not cause synchronization issues.

Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
especially with the edit awareness that Hazel provides. 
We leave further optimizations along these lines to future work.
Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.

>>>>

\section{Implementation}\label{sec:implementation}
Implementing livelits requires tight integration between a rich editor, 
a type checker, and a live evaluator 
capable of evaluating incomplete programs and gathering hole closures.

\subsection{Hazel}
Hazelnut Live is the foundation of the Hazel programming environment, 
and Hazel has support for all of the necessary mechanisms, so 
it was natural to choose Hazel for our primary implementation.
Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
% The core of Hazel is written in a pure functional style, so our implementation closely 
% follows the formalism.  
%  Hazel is a browser-based live functional programming environment 
% organized from the ground up around typed holes, with full .

The livelit definition mechanism described in Sec.~\ref{sec:livelit-definitions}
is implemented in Hazel, albeit without some of the syntactic sugar in Fig.~\ref{fig:color-impl}.
This, together with the fact that Hazel lacks a mature GUI widget libraries as of this writing, 
makes complex examples tedious to implement within Hazel, so we also 
added the ability to define livelits using JavaScript or OCaml. These are loaded when 
Hazel is compiled. As Hazel evolves, we expect to need to define such ``primitive'' livelits less 
frequently, using them mainly for livelits that would benefit from access to established JavaScript 
or OCaml libraries.

Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous. Hazel achieves this ``gap-free'' liveness guarantee by automatically inserting explicit holes as necessary 
while the user edits 
the program. Formally, Hazel is a type-aware structure editor \cite{Hazelnut}, rather than a text editor, 
although the developers are aiming to maintain a text-like experience (this effort is orthogonal to our own).

% The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
% implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
% to livelit definitions suitable for inclusion in the livelit context used in the semantics.

% This means that it is conceptually straightforward to hide the model and splice data and instead display 
% the computed view. However, there are a number of non-trivial layout challenges that come up, which we discuss
% in Sec.~\ref{sec:layout} below.

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-notation} 
is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
there are syntax errors.

\subsection{Layout}\label{sec:layout}
Whether implemented in a structure editor or a text editor, livelits present 
interesting layout challenges. 
Hazel uses an optimizing pretty printer based on the work of \citet{DBLP:journals/pacmpl/Bernardy17} to determine layout. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out either as inline livelits, like \li{\$slider},
which are one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width 
and a specified number of lines. % We omitted these specifications in Fig.~\ref{fig:color-impl}. 
% The pretty printer lays out multi-line livelits
% on their own line, as seen in Fig.~\ref{fig:grading}.

% We are currently exploring a more flexible system whereby 
% multi-line livelits can be named and applied within a larger expression, but
% the GUI is not displayed until the next line, even if there are intervening characters.
% This is inspired by the \verb|~| placeholder for large textual literals in the 
% Wyvern programming language \cite{TSLs}.

One might also considering a number of other layout 
options, e.g. inline-block literals \emph{a la} Wyvern \cite{TSLs}, pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
on a separate canvas or document while still formally being located within an underlying functional program.

This latter option 
would be particularly interesting for end-user programming scenarios: users with limited
programming experience 
could interact with a collection of livelits laid out separately in the popular ``dashboard'' style, 
without necessarily
even being aware that their interactions are actually edits to an underlying typed
functional program. 
% More experienced users on the team could then make programmatic use of the entered
% data and parameters in a seamless manner. We are also excited about the possibility of end users
% getting curious enough to ``View Source'' in this setting, and in so doing 
% being exposed to typed functional programming concepts gradually.

% The layout of splices within livelits is also an interesting problem, and our current approach, where 
% the provider must specify a fixed size for each splice, is not particularly flexible. 
% A more principled system for GUI layout that integrates with program pretty 
% printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
% livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
% a generalized formula bar in the editor user interface that multiple livelits could share.
% While splices can in principle contain other livelits, it is difficult for small splice editors to 
% visually accomodate large livelit views in practice. 
% Fortunately, variables can be used to support composition via substitution 
% rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.

% The Hazel pretty printer uses memoization and incremental diffs to maintain reasonable performance,
% and adding livelits to the system did not substantially interfere with the existing approach.
% For the relatively simple livelits that we have developed, performance has been adequate.
% However, more complex  livelits or livelits that update quickly 
% could easily cause problems with editor responsiveness.
% This could perhaps be resolved by performing livelit-related computations asynchronously. Care 
% would need to be taken to ensure that intervening edits do not cause synchronization issues.

% Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
% especially with the edit awareness that Hazel provides. 
% We leave further optimizations along these lines to future work.
% Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.

\subsection{Integration into Imperative Languages}
\label{sec:imperative-langs}
Our focus in this paper was on pure languages.
Side effects pose a challenge on two fronts.
If they occur in a livelit implementation,
then the state exists at edit-time. 
The most viable approach would be to retain the model-view-update
architecture and specify that any transient mutable state will be reset between updates.
If side effects occur in livelit expansions, 
then the closure collection mechanism requires more 
care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
However, these problems are surmountable, and we look forward to 
efforts to integrate livelits into imperative languages.

% The Elm architecture has proven practical for sophisticated user interfaces, 
% e.g. the Sketch-n-Sketch system \cite{sns-uist}, 
% and its purity addresses means there are no issues 
% related to initialization, maintenance, and persistence of mutable state
% by the editor.
% Imperative languages can express the same architecture (indeed, a close cousin 
% is at the heart of the popular React framework for Javascript).

### 2020-11-21 pldi21/intro.tex
Updates from Overleaf
updates

\subsection{Contributions}

We turn our attention in this paper to several fundamental technical
deficiencies that limit GUI providers and clients using these prior systems.
To address these, 
we introduce a system of \emph{live literals}, or \emph{livelits}, 
demonstrated in Fig.~\ref{fig:color}(b). 
Livelits are unique in achieving all of the following properties.
(We describe which subset of these properties 
are achieved by prior systems,
including those just mentioned, 
in Sec.~\ref{sec:related-work}.)

>>>>

\subsection{Contributions}
\label{sec:contributions}
We turn our attention in this paper to several fundamental technical
deficiencies that limit GUI providers and clients using these prior systems.
To address these, 
we introduce a system of \emph{live literals}, or \emph{livelits}, 
demonstrated in Fig.~\ref{fig:color}(b). 
Livelits are unique in achieving all of the following properties.
(We describe which subset of these properties 
are achieved by prior systems,
including those just mentioned, 
in Sec.~\ref{sec:related-work}.)

### 2020-11-21 pldi21/related-work.tex
updates

\section{Related Work}\label{sec:related-work}
As detailed in Sec.~\ref{sec:intro},
the \emph{Graphite} system introduced the idea of filling 
typed holes using a type-specific user interface \cite{Graphite}.
This prior work engaged in substantial qualitative evaluation, 
which due to the fundamental similarities between the two systems 
is as relevant to our design as theirs. However, the prior work 
left a number of core technical issues unresolved, 
as summarized in Sec.~\ref{sec:intro}. 
Livelits resolve these issues in large part by bringing together 
seemingly disparate ideas from several other pieces of recent work.

@@

This structural delimitation of different notations is reminiscent of work on 
\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused primarily on 
combining different notations, primarily textual, using structural delimiters inserted 
explicitly by the user using a special-purpose editor.,

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs \cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions can be 
included granularly in a livelit context and support decentralized extensibility.
Furthermore, livelit definitions are governed by a type and binding discipline that 
serves to ensure that a programmer unfamiliar with the semantics of an unfamiliar livelit
can still be sure that it is not operating outside of the bounds of the type system.
In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
can be approached: by inspecting its type signature and reasoning compositionally 
about arguments, i.e. splices.
Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
whereas other projectional editors either do not support evaluation of holes or do 
not use that information to provide feedback to the user.

Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
a great variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits are asked to create code where there would otherwise
be a hole based on interactions that the programmer has with a visual interface.

@@

The Vital programming environment for Haskell \cite{hanna2002interactive} supports type-specific stylesheets
that can be used to create custom user interfaces for both editing and displaying
values. The editors support splices, called \emph{cells}, that can contain 
Haskell code. Results are computed in a live manner, though there is no support
for evaluating incomplete programs. The visualizations themselves also cannot provide
live feedback. Moreover, the system does not enforce any hygiene or typing principles:
the user is entirely responsible for syntactic and semantic correctness, and the 
underlying semantics have not been formally defined.

>>>>

\section{Related Work}\label{sec:related-work}
As detailed in Sec.~\ref{sec:background},
the Graphite system developed the idea of filling 
typed holes using a type-specific user interface, and it was the starting 
point for our work \cite{Graphite}.
Subsequent work on \textbf{mage} further explores this design space \cite{DBLP:conf/uist/KeryRHMWP20}.
This prior work engaged in substantial qualitative evaluations, 
which due to the fundamental similarities between the two systems 
is as relevant to our design as theirs. However, the prior work 
left a number of core technical issues unresolved, 
as summarized in Sec.~\ref{sec:contributions}. 
Livelits resolve these issues in large part by bringing together 
seemingly disparate ideas from several other pieces of recent work.

@@

Recent work on a interactive visual macro system in Racket 
is quite similar in spirit to our work \cite{interactive-visual-syntax}.
However, it supports only a limited form of splicing via a pop up text editor
that does not support full compositionality as described here, where livelits 
can appear within other livelits. Splicing is not hygienic, nor is there any 
consideration of typing. Parameterization and partial application is also impossible, because invocation 
is via an editor command rather than a language-integrated mechanism. Finally,
and most critically, there is no support for liveness.

Our structural delimitation of splices is reminiscent of work on 
\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused on 
combining different notations, primarily textual, using structural delimiters inserted 
explicitly using a special-purpose editor.

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs 
\cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions can be 
included granularly in a livelit context and support decentralized extensibility.
Furthermore, livelit definitions are governed by a type and binding discipline that 
serves to ensure that a programmer unfamiliar with the semantics of an unfamiliar livelit
can still be sure that it is not operating outside of the bounds of the type system.
In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
can be approached: by inspecting its declarations and reasoning compositionally 
about arguments, i.e. splices and parameters.
Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
whereas many projectional editors do not support live evaluation or do 
not integrate liveness into projections.

A number of notebook systems, including Mathematica, Jupyter \cite{Guo13}, and others, 
support the insertion of simple widgets like sliders that respond live to changes.

Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
a variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits are asked to create code where there would otherwise
be a hole.

@@

The Vital programming environment for Haskell \cite{hanna2002interactive} 
supports type-specific stylesheets
that can be used to create custom user interfaces for both editing and displaying
values. The editors support splices, called \emph{cells}, that can contain 
Haskell code. Results are computed in a live manner, though there is no support
for evaluating incomplete programs. The visualizations themselves cannot provide
live feedback. Moreover, the system does not enforce any hygiene or typing principles:
the user is entirely responsible for syntactic and semantic correctness, and the 
underlying semantics have not been formally defined.
